# Handling social network data in R

This R notebook demonstrates some data handling procedures for relational (social network) data, especially data as generated by "name generator" questionnaires.

We start by loading some useful packages:

-   `igrapg` and `network` are packages specifically for handling network data;
-   `reshape2` and `tidyverse` are more general-purpose packages that included useful procedures.

**NOTE:** if you get any "there is no package called..." error messages, make sure you install these packages first using the `install.packages()` function (not included here).

```{r  message=F, warning =FALSE}
library(network)
library(igraph)
library(reshape2)
library(tidyverse)
```

## Getting network data into R: igraph

We start by reading some network data from disk (or from the internet, in this case). A simple and very common way to store data is as "comma-separated values" (CSV). Let's load some "toy data":

```{r}
url1 <- "https://github.com/rensec/sasr08/raw/main/g_adj_matrix_simple.csv"
g_matrix <- read.csv(file = url1, header = FALSE)
g_matrix
```

These data are stored as an *adjacency matrix*. While reading the data, R automatically created column names and loaded the data as a data frame:

```{r}
class(g_matrix)
```

For network analysis, this is often not practical; for example we may want to do matrix calculations. So instead, let's turn this into a matrix object:

```{r}
g_matrix <- as.matrix(g_matrix)
g_matrix
```

For clarity, we also add row- and column names, simply numbering them:

```{r}
rownames(g_matrix) <- 1:nrow(g_matrix)
colnames(g_matrix) <- 1:ncol(g_matrix)
g_matrix
```

**QUESTION:** How many nodes are included in this matrix, and how many edges are there between these nodes?

Typically, information about nodes and edges are stored in separate files. In our next step, we're reading an attribute of the nodes:

```{r}
g_nodes_age <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_nodes_age.csv")
g_nodes_age
```

We now have information about nodes and edges loaded in R . However, to R, these are just like *any* data; it doesn't "know" that these are network data. In order to be able to use specific SNA procedures, we first need to create *network objects*, which is what we'll do next.

**NOTE:** the data that we've loaded now just happened to be stored as an adjacency matrix; it is also possible to store network data on disk in other formats. We'll get to that later.

### Creating network objects in igraph

Network analysis packages such as `igraph` handle data efficiently by storing data in their own type of *objects*, which basically ensure that all procedures make sense for network data. A network object as used by `igraph` includes the information on both edges and nodes of a network in a single "box" (and potentially other information about the network too).

To create an `igraph` network object from our matrix object we run:

```{r}
g <- graph_from_adjacency_matrix(g_matrix)
class(g)
g
```

The output from running "g" already shows that this is now a network object (specifically: an object of the class "igraph", as shown by `class()`): R interprets the object as a network and shows us the edges in the network. This also means that other R functions - to the extent that they have network methods implemented - now automatically recognize `g` as a network object and will behave accordingly (note that this is a key feature of R as an "object-oriented language"). For example, we can now run the standard `plot()` function and it will create a network map:

```{r}
plot(g)
```

Now we can also add node attributes to the object (in this case we have only one):

```{r}
g <- set_vertex_attr(g, 
                     name = "age",
                     value = g_nodes_age$age)
```

### Creating network objects from adjacency list data

A common format for network data, especially if collected via survey methods, is the *adjacency list*. In this format, we have a row for each node, and columns that indicate the first, second, third...etc connection of each node. The file `g_adj_list.csv` is an example of this format (it is again our toy network as used in the previous excercises).

```{r}
g_adj_list <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_adj_list.csv")
g_adj_list
```

Note that:

-   The columns "friend1" and "friend2" could, for example, refer to "name generator" survey questions, where each node number in the column refers to a friend "nominated" by the node (respondent) in the corresponding row. In this case, the nodes can have at most two (outgoing) ties.

-   This format allows for easy inclusion of node attributes as well, such as, in this case, the column "age".

The most convenient way to create a network object from these data, is to first transform them into an *edge list*, as this is a format for which `igraph` has an import function, such that we can turn it into a network object.

To create an edge list, we first make the data *long*.

```{r}

g_elist_from_alist <- g_adj_list %>% 
  select(id, friend1:friend2) %>% #keep only the network variables (and id)
  melt( id.vars = "id") #from the reshape2 package
g_elist_from_alist

```

We could do the same using the `tidyr` package (automatically loaded with `tidyverse`:

```{r}
g_adj_list %>% # (we don't save it into an object as above this time, since we just want to show that the result is the same)
  select(id, friend1:friend2) %>% #keep only the network variables (and id)
  pivot_longer(c(friend1, friend2))
```

To create an edge list we subsequently drop all missing values on "value" and do some housekeeping:

```{r}
g_elist_from_alist <- g_elist_from_alist%>% 
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="id", to = "value", sourcevar= "variable") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
g_elist_from_alist
```

We can now interpret each {from,to} combination as a directed edge (tie) from respondent to alter; `sourcevar` is an *edge attribute* indicating where in the adjacency list columns the alter was mentioned. Strictly speaking you don't need the latter for a simple edge list, but it's useful to keep in case you want to reverse the procedure.

**QUESTION:** Besides useful more data manipulation procedures, keeping the information in `sourcevar` could also be important for more substantive reasons. Can you think of such a reason?

Now we can import this into a network object with an `igraph` function:

```{r}
g_from_elist <- graph_from_data_frame(g_elist_from_alist) #This is a proper igraph graph object
plot(g_from_elist)
```

Note however that we lost the isolate! To include the isolate, we need to feed the function also the list of *all* the nodes:

```{r}
nodelist <- select(g_adj_list,id,age)
g_from_elist <- graph_from_data_frame(g_elist_from_alist, vertices = nodelist) #This is a proper igraph graph object
plot(g_from_elist)
```

That's better! Note that besides the node IDs, we also include the age variable in the network object as a node attribute, such that we could use it in analyses based on the network object.

## The reverse direction: exporting from igraph objects

Handling your network data as igraph objects is useful for social network analysis, but sometimes we also want to transform our data back from network objects to "regular" matrix objects or data frames. We briefly cover some of these cases here.

### From igraph to adjacency matrix

```{r}
g_adj_matrix <- as_adjacency_matrix(g, sparse = FALSE)
g_adj_matrix

```

We here specify `sparse = FALSE` because by default, `as_adjacency_matrix()` will return a "sparse matrix" (just run `as_adjacenct_matrix(g)` on your R console and see what comes out), which is a somewhat more efficient way of handling matrices with many zeroes. To demonstrate that we can recreate our original adjacency matrix however, we don't want that here. To check whether the resulting matrix is indeed identical to what we originally read from disk we can do:

```{r}
all(g_adj_matrix == g_matrix)
```

### From igraph to edge list

```{r}

g_edgelist <- igraph::as_data_frame(g_from_elist, what = "edges") 
g_edgelist 

```

Note the use of the "namespace" "igraph::" here, to indicate that we need the igraph function here, not the function with the same name from the dplyr/tidyverse package

Note that now, once more, we have lost our isolated node 5! To get all nodes, we could simply run:

```{r}
igraph::as_data_frame(g_from_elist, what = "vertices") 
```

### Back to an adjacency list from igraph

```{r}
edgelist_2 <- igraph::as_data_frame(g_from_elist, what = "edges")
nodelist_2 <- igraph::as_data_frame(g_from_elist, what = "vertices")
# Note the use of the "namespace" "igraph::" here, to indicate that we need the igraph function here, not the function with the same name from the dplyr/tidyverse package

d <- edgelist_2 %>%
  rename(id = "from") %>% 
  pivot_wider( id_cols = id, names_from = sourcevar, values_from = to ) %>% 
  merge(nodelist_2, by.x = "id", by.y = "name", all.y = TRUE ) %>% 
  relocate(age, .after = id) %>% 
  type_convert() #as_data_frame returns characters; this transforms it back to numeric

d
```

...and we're back!

### From edge list to network package object

An alternative to `igraph` is the `network` package, which has its own type of network data object (as used by, for example, the `sna` and `ergm` packages), and its own set of functions for handling network data. Note that we've already loaded the `network` package at the start of the tutorial. We can import our data into a `network` object (using the edge list that we've created before) as follows:

```{r}
g_np <- as.network(g_elist_from_alist, matrix.type = "edgelist", vertices = nodelist)
g_np


```

```{r}
plot(graph_3)
```

We don't discuss `network` in-depth for now; it suffices to say that the choice between `igraph` and `network` will typically depend on the specific needs of your research project. Both packages include a large set of functions for handling network data, but as `igraph` is a bit more comprehensive in terms of functions for network *analysis*, it tends to be somewhat [more popular](https://www.r-bloggers.com/2016/01/r-graph-objects-igraph-vs-network/). There are also companion packages for `network` that allow various types of analyses, included in the `statnet` suite. Finally, for converting between `network`- and `igraph`- objects, look at `intergraph`.
