# Handling social network data in R {#tut_data_handling}

This tutorial demonstrates some data handling procedures for relational (social network) data. In particular, we'll learn:

-   How to read network data from a file and convert them into a 'network object' that R can work with, using the common network-oriented packages `igraph` and `network`;

-   How to move back and forth between different types of data formats for relational data;

-   How to perform basic data manipulation procedures in `igraph`;[^02_sasr08_tutorial_1_1_data_handling-1]

-   How to produce a basic visualization of a network.

[^02_sasr08_tutorial_1_1_data_handling-1]: `network` is also designed for handling network data, but we focus on `igraph` here for simplicity. The idea is that once you are familiar with `igraph`, using the procedures in `network` is relatively straightforward.

We start by loading some useful packages:

-   `igraph` and `network` are packages specifically for handling network data;
-   `reshape2` and `tidyverse` are more general-purpose packages that contain useful procedures.

**NOTE:** if you get any "there is no package called..." error messages, make sure you install these packages first using the `install.packages()` function (not included here).

```{r  message=F, warning =FALSE}
library(network)
library(igraph)
library(reshape2)
library(tidyverse)
```

## Getting network data into R: the igraph package

Network analysis packages such as `igraph` handle data efficiently by storing data in their own type of *objects*, which basically ensure that all procedures make sense for network data. Moreover, a network object as used by `igraph` allows to include the information on *both* edges and nodes of a network in a single "box" (and potentially other information about the network too).

The raw data that we use to create these network objects may come in different 'shapes', such as *adjacency matrices*, *edge lists*, or *adjacency lists*. In the following, we discuss how to turn data objects of these different shapes into `igraph` network objects.

### From adjacency matrix to igraph object

We start by reading some network data from a file (from the internet, in this case). A simple and very common way to store data is as "comma-separated values" (CSV). Let's load some "toy data":

```{r}
url1 <- "https://github.com/rensec/sasr08/raw/main/g_adj_matrix_simple.csv"
g_matrix <- read.csv(file = url1, header = FALSE)
g_matrix
```

These data are stored in the shape of an *adjacency matrix*, meaning that it is a square matrix in which both the rows and the columns represent nodes, and the values in the cells indicate the status of the edges between these nodes. While reading the data, R automatically created column names and loaded the data as a data frame:

```{r}
class(g_matrix)
```

For network analysis, this is often not practical; for example we may want to do matrix calculations. So instead, let's turn this into a matrix object:

```{r}
g_matrix <- as.matrix(g_matrix)
g_matrix
```

Now, conceptually, the data still have the *shape* of an adjacency matrix, but in R, it is now stored as an object of class 'matrix'.

For clarity, we also add row- and column names, simply numbering them:

```{r}
rownames(g_matrix) <- 1:nrow(g_matrix)
colnames(g_matrix) <- 1:ncol(g_matrix)
g_matrix
```

**QUESTION:** How many nodes are included in this matrix, and how many edges are there between these nodes?

Typically, information about nodes and edges are stored in separate files. In our next step, we're reading an attribute of the nodes:

```{r}
g_nodes_age <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_nodes_age.csv")
g_nodes_age
```

We now have information about nodes and edges loaded into R. However, to R, these are just like *any* data; it doesn't "know" that these are network data. In order to be able to use specific SNA procedures, we first need to create *network objects*, which is what we'll do next.

**NOTE:** the data that we've loaded now just happened to have the shape of an adjacency matrix; it is also possible to store network data to a file in other shapes. We'll get to that later.

To create an `igraph` network object from our matrix object we run:

```{r}
g <- graph_from_adjacency_matrix(g_matrix)
class(g)
g
```

The output from running "g" already shows that this is now a network object (specifically: an object of the class "igraph", as shown by `class()`): R interprets the object as a network and shows us the edges in the network. This also means that other R functions - to the extent that they have network methods implemented - now automatically recognize `g` as a network object and will behave accordingly (note that this is a key feature of R as an "object-oriented language"). For example, we can now run the standard `plot()` function and it will create a network map:

```{r}
plot(g)
```

Now we can also add node attributes to the object (in this case we have only one):

```{r}
g <- set_vertex_attr(g, 
                     name = "age",
                     value = g_nodes_age$age)
```

### From edge list to igraph object

A very common shape that network data come in is the *edge list*. This is a data matrix with usually just two columns, in which every row represents an edge between two nodes, listed in the columns.

The file 'g_elist.csv' contains an edge list for the same network 'g' that we saw above. Let's load the file into a data frame:

```{r}
g_elist <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_elist.csv")
g_elist
```

Here, each row represents an edge between the nodes listed in the columns 'from' and 'to'.

We can create an `igraph` network object from the edge list as follows:

```{r}
g_from_elist <- graph_from_data_frame(g_elist) 
plot(g_from_elist)
```

If you compare this network with the one pictured earlier, you'll notice an important difference: we lost the isolate! This is an important limitation of edge lists: due to their nature, they do not include isolates.

To include the isolate in the network object, we need to feed the function also the list of *all* the nodes. Fortunately we already have the complete list of nodes, include in our data object with node attributes (g_nodes_age). We can include the list of nodes when we import the edge list into the `igraph` object:

```{r}
g_from_elist <- graph_from_data_frame(g_elist, vertices = g_nodes_age) 
plot(g_from_elist)
```

Now the isolate is included as it should. Thus, when importing edge lists, it is important to always include a complete node list as well (unless you are certain that the edge list includes all nodes)!

Note also that now we have included the node attribute 'age' already with the node list when we created the network object. Thus, we so far have seen two ways to add node attributes to a network object:

1.  Include them directly with a node list when you first create the network object, as specified in the 'vertices' argument of a function like `graph_from_data_frame()`;

2.  Add them afterwards using `set_vertex_attr()`.

**Assignment:** from the code snippets above, construct a complete "pipeline" to go from the edge list to the igraph object. Start from `g_elist`. Plot the resulting network. Store the code somewhere for later use.

### From adjacency list to igraph object

Another common shape for network data, especially if collected via survey methods, is the *adjacency list*. In this shape, we have a row for each node, and columns that indicate the first, second, third... etc. connection of each node. The file `g_adj_list.csv` is an example of this format (it is again our toy network as used in the previous excercises).

```{r}
g_adj_list <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_adj_list.csv")
g_adj_list
```

Note that:

-   The columns "friend1" and "friend2" could, for example, refer to "name generator" survey questions, where each node number in the column refers to a friend "nominated" by the node (respondent) in the corresponding row. In this case, the nodes can have at most two (outgoing) ties.

-   This format allows for easy inclusion of node attributes as well, such as, in this case, the column "age".

Unfortunately, `igraph` does not have a function to directly import adjacency lists.[^02_sasr08_tutorial_1_1_data_handling-2] The most convenient way to create a network object from these data, is to first transform them into an *edge list*, as this is a format for which `igraph` has an import function, such that we can turn it into a network object.

[^02_sasr08_tutorial_1_1_data_handling-2]: Actually there is an igraph function 'graph_from_adj_list()', but this expects a different data structure than what we have here.

To create an edge list, we first make the data *long*.

```{r}
g_elist_from_alist <- g_adj_list
g_elist_from_alist <-  select(g_elist_from_alist, id, friend1:friend2) # keep only the network variables (and id)
g_elist_from_alist <-  melt(g_elist_from_alist, id.vars = "id") # from the reshape2 package
g_elist_from_alist

```

To streamline this code a little bit, we could use the "pipe" operator (`%>%`) which comes with the `tidyverse` package. This does exactly the same, but allows us to write the code more compactly. In RStudio, you can insert it easily using the CTRL-SHIFT-M keyboard shortcut (in Windows).

```{r}

g_elist_from_alist <- g_adj_list %>% 
  select(id, friend1:friend2) %>% # keep only the network variables (and id)
  melt( id.vars = "id") # from the reshape2 package

```

We could achieve the same result using the `tidyr` package (automatically loaded with `tidyverse`:

```{r}
g_adj_list %>% # (we don't save it into an object as above this time, since we just want to show that the result is the same)
  select(id, friend1:friend2) %>% #keep only the network variables (and id)
  pivot_longer(c(friend1, friend2))
```

To create an edge list we subsequently drop all missing values on "value" and do some housekeeping:

```{r}
g_elist_from_alist <- g_elist_from_alist%>% 
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="id", to = "value", sourcevar= "variable") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
g_elist_from_alist
```

We can now, once again, interpret each {from,to} combination as a directed edge (tie) from respondent to alter; in addition, `sourcevar` is an *edge attribute* indicating where in the adjacency list columns the alter was mentioned. Strictly speaking you don't need the latter for a simple edge list, but it's useful to keep in case you want to reverse the procedure.

**QUESTION:** Besides being useful for data manipulation procedures, keeping the information in `sourcevar` could also be important for more substantive reasons. Can you think of such a reason?

Now we can import this into a network object with an `igraph` function:

```{r}
g_from_alist <- graph_from_data_frame(g_elist_from_alist) #This is a proper igraph graph object
plot(g_from_alist)
```

Note however, once more, that we lost the isolate! To include the isolate, we need to feed the function also the list of *all* the nodes, as we did before. However, in this case, we can also create the node list from the adjacency list data frame itself:

```{r}
nodelist <- select(g_adj_list,id,age)
g_from_alist <- graph_from_data_frame(g_elist_from_alist, vertices = nodelist) #This is a proper igraph graph object
plot(g_from_alist)
```

That's better! Note that besides the node IDs, we also include the age variable in the network object as a node attribute, such that we could use it in analyses based on the network object.

**Assignment:** from the code snippets above, construct a complete "pipeline" to go from the adjacency list to the igraph object. Start from `g_adj_list`. Plot the resulting network. Store the code for later use.

## The reverse direction: exporting from igraph objects

Handling your network data as igraph objects is useful for social network analysis, but sometimes we also want to transform our data back from network objects to "regular" matrix objects or data frames. We briefly cover some of these cases here. Typically, we can use dedicated `igraph` functions for this purpose.

### From igraph to adjacency matrix

```{r}
g_adj_matrix <- as_adjacency_matrix(g, sparse = FALSE)
g_adj_matrix

```

We here specify `sparse = FALSE` because by default, `as_adjacency_matrix()` will return a "sparse matrix" (just run `as_adjacenct_matrix(g)` on your R console and see what comes out), which is a somewhat more efficient way of handling matrices with many zeroes. To demonstrate that we can recreate our original adjacency matrix however, we don't want that here. To check whether the resulting matrix is indeed identical to what we originally read from a file we can do:

```{r}
all(g_adj_matrix == g_matrix)
```

### From igraph to edge list

```{r}

g_edgelist <- igraph::as_data_frame(g_from_elist, what = "edges") 
g_edgelist 

```

Note the use of the "namespace" "igraph::" here, to indicate that we need the igraph function here, not the function with the same name from the dplyr/tidyverse package

Note that now, once more, we have lost our isolated node 5, as it is not included in the edge list! As before, we need a list of nodes in to our edge list to have a complete overview of the network, including any isolates. To get the nodes list, we could simply run:

```{r}
igraph::as_data_frame(g_from_elist, what = "vertices") 
```

### From igraph to adjacency list

```{r}
edgelist_2 <- igraph::as_data_frame(g_from_alist, what = "edges")
nodelist_2 <- igraph::as_data_frame(g_from_alist, what = "vertices")
# Note the use of the "namespace" "igraph::" here, to indicate that we need the igraph function here, not the function with the same name from the dplyr/tidyverse package

d <- edgelist_2 %>%
  rename(id = "from") %>% 
  pivot_wider( id_cols = id, names_from = sourcevar, values_from = to ) %>% 
  merge(nodelist_2, by.x = "id", by.y = "name", all.y = TRUE ) %>% 
  relocate(age, .after = id) %>% 
  type_convert() #as_data_frame returns characters; this transforms it back to numeric

d
```

...and we're back!

## Getting network data into R: the network package

An alternative to `igraph` is the `network` package, which has its own type of network data object (as used by, for example, the `sna` and `ergm` packages), and its own set of functions for handling network data. Note that we've already loaded the `network` package at the start of the tutorial. We can import our data into a `network` object (using the edge list that we've created before) as follows:

```{r}
g_np <- as.network(g_elist_from_alist, matrix.type = "edgelist", vertices = nodelist)
g_np


```

```{r}
plot(g_np)
```

We don't discuss `network` in-depth for now; it suffices to say that the choice between `igraph` and `network` will typically depend on the specific needs of your research project. Both packages include a large set of functions for handling network data, but as `igraph` is a bit more comprehensive in terms of functions for network *analysis*, it tends to be somewhat [more popular](https://www.r-bloggers.com/2016/01/r-graph-objects-igraph-vs-network/). There are also companion packages for `network` that allow various types of analyses, included in the `statnet` suite. Finally, for converting between `network`- and `igraph`- objects, look at `intergraph`.

## Modifying networks in igraph

We now turn back to `igraph`. Once we have our data wrapped into a network object, we can use `igraph` functions to make changes to the data. For example, we may remove a node:

```{r}
g_mod <- delete_vertices(g,2)
plot(g_mod)
```

This simple operation illustrates the power of handling networks as network objects: teh `igraph` function `delete_vertices()` "understands" that removing a node from a network logically implies that also the edges connected to this node should be removed. If you would have to do this with, say, a raw edge list or adjacency list, it would be much more cumbersome!

The same function also takes a vector of node IDs. For example, to remove all nodes with age = 21, we could do:

```{r}
g_mod <- delete_vertices(g,which(V(g)$age == 21))
plot(g_mod)
```

In the above code, we use the `igraph` function `V()` to get all vertices of the graph.

**QUESTION:**: Write the code to remove all nodes from `g` for which age is missing.

We have already seen the use of `set_vertex_attr()` to add node attributes, of `V()` to get all nodes. As an alternative to `set_vertex_attr()`, we can also use `V()` for the same purpose:

```{r}
V(g)$gender <- c("male", "female", "female", "other", "male") 
g
```

Similarly, we can use `E()` to access, add and modify edge attributes.

We can also use functions that modify the entire graph. For example, we may use `reverse()` to reverse the direction of all edges:

```{r}
g_rev <- reverse_edges(g)
plot(g_rev)
```

We can use `union()` to combine the edges of two graphs:

```{r}
g_union <- igraph::union(g, g_rev)
```

***QUESTION***: Before you run `plot(g_union)`, think about what the resulting should look like. Thus, what could you use the combination of `reverse()` and `union()` for?

A similar result could be produced with:

```{r}
g_undir <- as_undirected(g)
plot(g_undir)
```

Realize however that conceptually, g_union and g_undir are different!

***QUESTION:*** What's the difference?

For more information on all the graph manipulation functions included in `igraph`, see the [package's website](https://r.igraph.org/articles/igraph.html) and, of course, the help pages in R.
