# Basic network analysis in R

In this tutorial we'll use the `igraph` package to do some basic descriptive social network analysis. Obviously, we can only scratch the surface here, and many research project will require different analyses. Also, we do not cover visualization specifically. 

As an example network, we'll use the famous "karate club" data studied by Zachary (1977), included with `igraph`.

```{r}
library(igraph)
z <- make_graph("Zachary")
z
```

```{r}
plot(z)
```


This looks like an undirected network, which we can verify using:

```{r}
is_directed(z)
```


## Network level: a "five number summary"

A good starting point of any network analysis project is to look at some basic properties of the graph *as a whole*. While there are many potentially interesting properties, we'll focus here on the the "five number summary" as suggested by [Luke (2015)](https://link.springer.com/book/10.1007/978-3-319-23883-8).

### Size

Probably the most basic property of a network is its *size*, that is, the number of vertices and the number of edges. While `igraph` already reported them above (if you knew where to look), we can ask for these numbers specifically:

```{r}
ecount(z)
```
```{r}
vcount(z)
```


### Density

The *density* of a network is the ratio of existing edges and the maximum possible number of edges that could exist, given the number of vertices. We can let R calculate it for us:

```{r}
edge_density(z)
```

Obviously, we could already have calculated this ourselves, given that we already knew the number of edges and the number of vertices.

**QUESTION**: How would you calculate density using `ecount()` and `vcount()`?

If desired, we can also add such network-level results to the network object, for later use:

```{r}
z$density <- edge_density(z)
z
```
'Density' is now listed as one of the attributes of the object, where 'g/n' indicates that it is a graph-level attribute. 

### Components

The number of components is the number of unconnected parts of the network (which may be parts consisting of one node, that is, isolates). The number of components in this network is, quite trivially, just one, but the way to get it is:
```{r}
count_components(z)
```


### Diameter (and distances)
The diameter of a network is the "longest shortest path" in the network
```{r}
diameter(z)
```

A related an much-studied property is the *average* shortest path length, that is, the average of all the shortest path over all pairs of vertices in the network. For example, this is the key indicator in the "small world phenomenon".  

```{r}
mean_distance(z)
```



### Clustering Coefficient

Clustering, or transitivity, relates to the extent to which triangles tend to be closed in the network, or put differently, to what extent neighbors of nodes tend to be connected themselves. There are many ways to quantify this tendency (which also differ for directed an undirected networks), and the `transitivity()` function covers many of them. Note that in the literature and elsewhere in the field, terms like 'transitivity' and 'clustering'  are often used quite loosely, so it is always wise to look closely at the formal specifications (if provided) to know what is meant in a specific use case.

We here compute what is most commonly known as the 'clustering coefficient'. This computes, for each vertex, the proportion of the potential ties between the vertex' neighbors that actually exist (in the ego networks literature, this is referred to as local density), and averages this over all vertices.  

```{r}
transitivity(z, type="average")
```

Interestingly, the specification of 'average' for the 'type' parameter is not explained in the `igraph` documentation, but this is what it does. 

## Individual level: centrality

### Degree centrality

```{r}
degree(z)
```

```{r}
V(z)$degree <- degree(z)
```

Alternatively, we could have used `set_vertex_attr()`, which would work better in a "pipeline" workflow (using `%>%`).


### Betweenness centrality

### Closeness centrality