# Transform the Knecht data to dyads-as-cases

## Prliminaries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(reshape2)
library(igraph)
```


```{r}
PupilsWaveV <- read_dta(file = "PupilsWaveV.dta")
```


## Data preparation



We start with two classes.

```{r}
df <- PupilsWaveV %>% 
  filter(schoolnr == "12b" | schoolnr == "10a")  # keep two classes
table(df$schoolnr) # check the number of cases)
```

We now create a dyads-as-cases dataset.

First we define the function to create such a dataset for one class, which we will then apply to each class separately.

*NOTE:*: This function assumes that we have no "missing egos" (see the tutorial on "the case of the missing ego" for a solution).

```{r}
create_dyads <- function(subnet){
  # Create the edgelist
  dyads <- subnet %>% 
  select(namenr, friend1:friend2) %>% 
  melt(id.vars = "namenr") %>% #make long
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="namenr", to = "value", sourcevar= "variable") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
  
  dyads$connected <- 1 # add a variable indicating that they are connected
  
  # Create a list of all possible dyads
  dyads <- dyads %>% 
    full_join(expand.grid(from = subnet$namenr, to = subnet$namenr), by = c("from", "to")) %>% # create a list of all combinations of nodes, and match this to whether they are connected
    mutate(connected = ifelse(is.na(connected), 0, connected)) %>%  # if there is no connection, set connected to 0
    filter(from != to) %>% # remove the diagonal
    select(from, to, connected) # keep only from, to and connected
  
  # Add the group identifier
  dyads$schoolnr <- unique(subnet$schoolnr)
  dyads <- select(dyads, schoolnr, from, to, connected) # and move it to the front for convenience
  
  
return(dyads)
}
```

We can now apply this function to each subset of the data frame, and then bind the results together.

```{r}
dyads <- df %>% 
  group_split(schoolnr) %>% # Split the dataframe into a list of dataframes, one for each value of schoolnr
  map(create_dyads) %>%  # apply the function to each subset
  list_rbind() # bind the results together
```

Let's check if the number of cases is correct. 
```{r}
nclass <- table(df$schoolnr) # number of cases in each class
nclass
```

The total number of cases in `dyads` should be: 
```{r}
nclass[[1]]*(nclass[[1]]-1) + nclass[[2]]*(nclass[[2]]-1)
```

Important: check for yourself that you understand: 1) what this code does, and 2) why this should be the total number of dyads. Now let's check: 

```{r}
nclass[[1]]*(nclass[[1]]-1) + nclass[[2]]*(nclass[[2]]-1) == nrow(dyads)
```

## Creating identifyers
For multilevel analysis, we will need four identifiers: one for the class, one for the ego (from), one for the alter (to), and one for the dyad. We already have the class id (schoolnr). We can create the other three as follows.

For the individual identifiers, we can simply combine the class id with the from and to variables.  
```{r}
dyads <- dyads %>%
  mutate(from_id = paste0(schoolnr, "_",as.character(to)),
         to_id = paste0(schoolnr, "_",as.character(from))
         )
```

For the dyad identifier, we need an id that is the same for both directions of the dyad (i.e., from A to B should have the same id as from B to A). We can achieve this by sorting the from and to variables, and then combining them.
```{r}
dyads <- dyads %>%
  mutate(pair_id = paste0(schoolnr, "_", pmin(from,
                                            to), "_", pmax(from, to)))

```

## Adding individual variables

We can add the gender for each pupil by merging on the original data frame. We need to do this twice: once for the from variable, and once for the to variable. Because we have no unique pupil identifier in the PupilsWaveV data, we will merge on the combination of schoolnr and namenr.

```{r}
dyads <- dyads %>%
  left_join(df %>% select(schoolnr, namenr, sex), by = c("schoolnr", "from" = "namenr")) %>% 
  rename(sex_from = sex) %>% 
  left_join(df %>% select(schoolnr, namenr, sex), by = c("schoolnr", "to" = "namenr")) %>%
  rename(sex_to = sex) %>% 
  mutate(sex_to = sex_to-1,
         sex_from = sex_from-1) # recode to 0 and 1 for easier interpretation
```


Create a variable for gender similarity, which is 1 if both pupils have the same gender. We do so by simply taking the absolute difference between the genders of both pupils in a dyads.
We then subtract this from 1, so that a value of 1 indicates that both pupils have the same gender.

```{r}
dyads  <- dyads %>%
  mutate(sex_similarity = 1-abs(sex_to-sex_from))
  
```

Finally, we save the data:

```{r}
write_csv(dyads, "dyads_knecht.csv")
```
