---
title: "Importing networks from the Knecht data, revisited"
author: "Rense Corten"
date:  "`r Sys.Date()`"
output: html_notebook
---

# Introduction

Creating network objects based on name generator questions is less straightforward than you might think. The method we used before actually has a flaw; here we provide a corrected version. 

```{r}
library(haven) #For reading Stata files
library(network) #For data handling of network data 
library(tidyverse) # because Tidyverse!
```

# Preparation

Get the data:

```{r}
PupilsWaveV <- read_dta(file = "PupilsWaveV.dta")
class12b <- PupilsWaveV  %>% 
  filter(schoolnr == "12b")  # keep only one class

```


Select the friends network:

```{r}
m <- class12b %>% 
  select(namenr, friend1:friend12)  # select only the " best friends"  network

```

Let's artificially add an isolate, for fun:

```{r}
iso <- m %>% 
  filter(namenr == 15) %>% 
  mutate(namenr = 27)
m <- rbind(m,iso)

length(m$namenr)

```
We now have 27 nodes. Node 27 is an isolate.

# The old method

Create an edgelist:
```{r}
el <- as.matrix(m)  # so that the next line works as intended; cbind() works in specific ways on matrix objects, which we exploit here.
el<- cbind(el[,1],c(el[,-1])) %>% 
  as_tibble() %>% # So that we can easily filter 
  filter(!is.na(V2)) # Now it is an edge list

```

Create a network object :
```{r}
g1 <- as.network.matrix(el,matrix.type = "edgelist") #from network package  
g1
```
Now here lies the problem: node 27 is gone!

# The improved method

We also create a list of nodes:

```{r}
nl <- m %>% 
  select(namenr) %>% 
  as.matrix() %>%  # Going to matrix and back to df removes the column attributes (such as var lables), which are a problem for the network package
  as.data.frame()

```

Now import again, but specify the list of nodes explicitly, as allowed by the as.network() function:

```{r}
g2 <- as.network(el, matrix.type = "edgelist", vertices = nl)
g2
```


Now we have 27 nodes, including the isolate. Using this method, it is also possible to include other edge attributes with the node list (see ?as.network()).