---
title: "SaSR 8 Tutorial 1.1"
subtitle: "Excercises for chapter 5: data management"
author: "Rense Corten"
output: html_notebook
---

# Introduction

This document contains a run-through of selection the exercises with Chapter 5 of "Analyzing Social Networks Using R" by Borgatti, Everett, Johnson, and Agneesens (hereafter ANSUR by BEJA). The original exercises can be found here: <https://www.analyzingsocialnetworksusingr.com/home/5-data-management>. We skip some exercises, and also correct a few minor errors in the original material.

This document was written as an [R Markdown](http://rmarkdown.rstudio.com) Notebook. If you are reading a html or pdf version of this, you can just keep reading, and if you want to run any of the code yourself, you need to copy and paste the code to your own R-script.

If you are reading this in the .rmd format: Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*. When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

# Installation

The book relies almost exclusively on the custom-made R package xUCINET. To install the xUCINET package, follow the instructions on <https://www.analyzingsocialnetworksusingr.com/xucinet> (note that the instructions on p.91 in the book are incorrect; install.packages() does not work), but with the following exceptions, but in step 3, install the packages "blockmodeling" and "FactoMineR".

Afterwards, you can load the package in R using library(). We also load tidyverse for convenience.

```{r, message= FALSE, warning = FALSE}
library(xUCINET)
library(tidyverse)
```

# Chapter 5: Data management

## 5.2: Checking for mistakes

In this first step, we'll have a look at wahat network data may look like in R, and how to fix some typical mistakes. First we create some input data, in the form of an adjacency matrix:

```{r}
adj_matrix_1 <- data.frame(A = c("ID", "Andrea", "Elsa", "Rafael", "Vera", "Peter", "Jared"),
                           B = c("Andrea", "0", "1", "0","0", "1","1"),
                           C = c("Elisa", "1", "1", "1","0", "0","0"),
                           D = c("Rafael", "1", "1", "0","1", "0","0"),
                           E = c("Vera", "0", "1", "1","0", "0","0"),
                           F = c("Gert", "0", "1", "0","1", "0","0"),
                           G = c("Peter", "1", "1", "2","1", "0",""),
                           H = c("Jared", "0", "0", "0","0", "1","1")
)
adj_matrix_1
```

Ideally this should be a square matrix, with rows and columns for every person. However, it appears Gert is missing from the rows. So let's add a new row for Gert:

```{r}
adj_matrix_2 <- data.frame(A = c("Gert"),
                           B = c(""),
                           C = c(""),
                           D = c(""),
                           E = c(""),
                           F = c(""),
                           G = c(""),
                           H = c("")
) %>% 
  rbind(adj_matrix_1)

adj_matrix_2 <- adj_matrix_2[c(2:6,1,7:8),] #put Gert in the right place.
rownames(adj_matrix_2) <- 1:nrow(adj_matrix_2) # reset the row names
adj_matrix_2
```

Also, some cells are empty. After verifying that this is not a coding or transcription mistake (which we assume now is not the case, we may want to designate these cells as missing explicitly, for example using the value -999.

```{r}
adj_matrix_2[adj_matrix_2 == ""] <- "-999"


```

Finally, we note that the rows mention "Elsa" while the columns mention "Elisa". Assuming that we concluded that "Elsa" is correct:

```{r}
adj_matrix_2[adj_matrix_2 == "Elisa"] <- "Elsa"
```

Done!

```{r}
adj_matrix_2
```

Wait! There is one more (potential) mistake. Can you spot it, and do you know how to solve it?

## 5.3: Read in and create a data project file

First we manually create the input data again:

```{r}


adj_matrix_friends <- data.frame(
                           A = c("ID", "Ana", "Ella", "Isra", "Petr", "Sam"),
                           B = c("Ana",  "0", "1", "3","2", "3"),
                           C = c("Ella", "2", "0", "1","1", "2"),
                           D = c("Isra", "1", "3", "0","2", "1"),
                           E = c("Petr", "3", "1", "2","0", "1"),
                           F = c("Sam",  "2", "3", "1","1", "0")
)


adj_matrix_communication <- data.frame(
                           A = c("ID", "Ana", "Ella", "Isra", "Petr", "Sam"),
                           B = c("Ana",  "0", "1", "0","1", "1"),
                           C = c("Ella", "1", "0", "0","1", "0"),
                           D = c("Isra", "0", "0", "0","1", "0"),
                           E = c("Petr", "1", "1", "1","0", "1"),
                           F = c("Sam",  "1", "0", "0","1", "0")
)

nodes_attributes <- data.frame(
                           A = c("ID",  "Ana", "Ella", "Isra","Petr","Sam"),
                           B = c("Age",   "23",  "67",   "45",  "28",  "33"),
                           C = c("Gender","1",   "1",    "1",   "2",   "2"),
                           D = c("Music", "Reggae", "Rock", "Pop","Reggae", "Pop")
  
)

```

And then save them as cvs-files:

```{r}
write.table(adj_matrix_friends,file = "PartyOf5_Friendship1.csv", row.names=FALSE, col.names = FALSE, sep = ",")
write.table(adj_matrix_communication,file = "PartyOf5_Communication1.csv", row.names=FALSE, col.names = FALSE, sep = ",")
write.table(nodes_attributes,file = "PartyOf5_Attributes1.csv", row.names=FALSE, col.names = FALSE, sep = ",")


```

Now we can load them as a network object:

```{r}

SmallProject5<-xCreateProject(GeneralDescription="This is a small study among 5 people.",
                              NetworkName="Friendship1",
                              NETFILE="PartyOf5_Friendship1.csv",
                              FileType="csv",
                              InFormatType="AdjMat",
                              NetworkDescription="Friendship networks among 5 people with 3=Good friend, 2=Friend and 1=Acquaintance",
                              Mode = c("People"),
                              Directed=TRUE, 
                              Loops=FALSE,
                              Values="Ordinal",
                              Class="matrix",
                              References="No references")

```

Note that although this is a lot of syntax, it writes out a lot of defaults.

Let's add the attribute data:

```{r}
SmallProject5 <- xAddAttributesToProject(ProjectName = SmallProject5,
                                ATTFILE1="PartyOf5_Attributes1.csv",
                                FileType = "csv",
                                Mode = c("People"),
                                AttributesDescription=c("Age in years","Gender (1=Male, 2=Female)","Preferred music")
                                )
```

```{r}
SmallProject5<-xAddToProject(ProjectName=SmallProject5,
                             NetworkName="Communication9",
                             NETFILE1="PartyOf5_Communication1.csv",
                             FileType="csv",
                             NetworkDescription="Communication network",
                             Mode=c("People"),
                             Directed=FALSE,
                             Loops=FALSE,
                             Values="Binary",
                             Class="matrix")
```


## 5.5. Transposing, symmetrizing and combining

### 5.5.1: Transposing

```{r}

Baker_Journals$Citations #Note: the website mistakenly mentions "Cocitations"
xTranspose(Baker_Journals$Citations)

```

### 5.5.2: Symmetrizing

We can symmetrize in different ways.

```{r}
xSymmetrize(Baker_Journals$Citations)
```

```{r}
xSymmetrize(Baker_Journals$Citations, Type = "Max")
```

```{r}
xSymmetrize(Baker_Journals$Citations, Type = "Av")
```

```{r}
xSymmetrize(Baker_Journals$Citations, Type = "Sum")
```

By the way, sometimes error messages in R (or perhaps this particular package) can be rather unhelpful. Note what happens when we mistype the 'Type="Sum"' option, instead writing 'Type = "sum"':

```{r}
xSymmetrize(Baker_Journals$Citations, Type = "sum")
```

We can't be expected to know what 'OUTPUT1' refers to, as it is not something in our code, and nothing here points to the misspecification of the Type option. The lesson here is to always closely crosscheck your code with the syntax specification in the help file whenever you get errors.

## 5.5.4 Combining

Again, there are different ways of combining. Here we try "minimum" and "sum".

**QUESTION**: what do these two options *substantively* represent?

```{r}
xCombineTies(Hawthorne_BankWiring$Conflict, Hawthorne_BankWiring$Friendship, Type = "Min")
```

```{r}
xCombineTies(Hawthorne_BankWiring$Conflict, Hawthorne_BankWiring$Friendship, Type = "Sum")
```

**QUESTION:** Given that we are combining "conflict" and "friendship" ties here, perhaps the two examples above are substantively not the most sensible ways to combine the two. What do you think would be a more useful way to combine these two networks? And does xCombineties() provide a way of doing that (tip: check the help file) ?

## 5.11 Saving and exporting

Sometimes we want to save a network dataset. There are several ways of doing so. First, we can simply save an R object as an R data file, which we can then load at a later point. Note that unless you specify the path as well, save() and load() use your current working directory, which you can set with setwd(). It is a good habit to avoid using full paths in code, and instead point your software (RStudio in this case) to the correct folder before you run the code. This makes your code much more portable between different devices and environments.

```{r}
HTnet <- Hawthorne_BankWiring # This step is not strictly necessary but useful for demonstration purposes
getwd() # shows your current working directory; change with setwd() if you are unhappy with it
save(HTnet, file="HT.RData")
rm(HTnet)
load("HT.RData")
```

The .RData format, however, only works with R. Sometimes we want to export data that we can also use in other software. Then, .csv ("comma-separated values") is an often-used format. But because csv-files are simply text files, we cannot save an entire multidimensional R-object (containing multiple networks, attributes, and project info, etc.) in a single csv file, but only one "dimension" at a time. Here, we choose the friendship network from the Hawthorne data set:

```{r}
write.csv(Hawthorne_BankWiring$Friendship, file = "HTnet_friends.csv")
HTnet_friends <- read.csv(file = "HTnet_friends.csv")
HTnet_friends
```

Note that this adjacency matrix, to R, is just a data frame, not a part of a network object. If we were to export the entire Hawthorne data object as a set of csv files, and import it back into R one by one, we'd have to piece back together the network object first if we wanted to work with it in xUCINET().

Finally, we may want to work with network data using different network analysis packages, such as [sna](https://cran.r-project.org/web/packages/sna/index.html) (which is part of the [statnet](https://statnet.org/) suite) or [igraph](https://r.igraph.org/).

```{r}
library(sna)

HBF.n<-as.network(Hawthorne_BankWiring$Friendship, directed = TRUE)
HBF.n
```

```{r}
library(igraph)
HBF.i<-graph_from_adjacency_matrix(Hawthorne_BankWiring$Friendship,
  mode = c("undirected"), diag = FALSE)
HBF.i
```
