# Solutions to Tutorial 4 - Spatial Models in R 

## Preliminaries

Load different packages:

```{r }
library(haven)
library(tidyverse)
library(igraph)
library(lme4)
library(spdep)
library(spatialreg)
```

## Open the dataset

Get Knecht data:

```{r }
PupilsWaveV <- read_dta(file = "PupilsWaveV.dta")
```

## Linear regression

Clean variables:

```{r }
PupilsWaveV$acthomew <- 5-PupilsWaveV$acthomew
PupilsWaveV$sex <- PupilsWaveV$sex-1
PupilsWaveV$sex <- 1-PupilsWaveV$sex
PupilsWaveV$acthomew[PupilsWaveV$acthomew == 9] <- NA
```

Linear regression:

```{r }
olsreg <- lm(acthomew ~ sex + age, data=PupilsWaveV)
summary(olsreg)
```

We see that girls do more homework than boys, but this is not statistically significant.

## Multilevel model

```{r }
ml <- lmer(acthomew ~ sex + age + (1 | schoolnr), PupilsWaveV)
summary(ml)
```

We again see that girls do more homework than boys, but this is not statistically significant.

## Adjacency matrix

Keep only the class 12b:

```{r }
class12b <- PupilsWaveV %>% 
  filter(schoolnr == "12b")  # keep only one class
```

Fetch the spatial weight matrix:

```{r }
friendsadjlist <- class12b %>%
  select(namenr, friend1:friend12)
class(friendsadjlist)
```

It is currently in the adjacency list format.
We first transform these data into edge list format:

```{r }
friendsedgelist <- class12b %>% 
  select(namenr, friend1:friend12) %>%   # select only the " best friends"  network
  pivot_longer(c(friend1:friend12)) %>% 
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="namenr", to = "value", sourcevar= "name") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
```

We lost the potential isolates (observation 15).
We also would like information on the nodes:

```{r }
friendsnodelist <- class12b %>% 
  select(namenr,sex,age,acthomew) %>% 
  rename(name = "namenr")
```

Transform to igraph object:

```{r }
friendsgraph <- graph_from_data_frame(friendsedgelist, vertices = friendsnodelist)
class(friendsgraph)
```

Plot the friends network:

```{r }
plot(friendsgraph)
layout <- layout_with_kk(friendsgraph)
plot(friendsgraph, layout = layout, edge.arrow.size = .5)
```

Transform to adjacency matrix:

```{r }
friendsadjmatrix <- as_adjacency_matrix(friendsgraph, sparse = FALSE)
class(friendsadjmatrix)
```

## Transform to a weights list object

```{r }
friends <- mat2listw(friendsadjmatrix, row.names = NULL, zero.policy =TRUE, style="W")
class(friends)
```

Style W means row-standardized.

## Moran's I test and LM tests

Moran's I test:

```{r }
moran.test(class12b$acthomew, friends, zero.policy = TRUE)
```

We set zero.policy to TRUE because we have isolates.
Moran's I test does not reject the null hypothesis, suggesting a lack of dependence in the data.
However, we only have 26 observations, making it very difficult to reject any null hypothesis.
Thus, we will ignore this.

Lagrange multiplier tests for spatial lag and spatial error dependencies:

```{r }
olsregnew <- lm(acthomew ~ sex + age, data=friendsnodelist)
summary(olsregnew)
lm.RStests(olsregnew, friends, test=c("LMlag", "LMerr"), zero.policy = TRUE)
```

We reject the null hypothesis for the LMlag, but not for the LMerr.
Hence, we should use the spatial lag model. But we will use both in this exercise.

## Spatial lag model

```{r }
slm <- lagsarlm(acthomew ~ sex + age, data = friendsnodelist, friends, zero.policy = TRUE, Durbin=FALSE)
summary(slm)
```

We only interpret the coefficients because of the low sample size, we will ignore the significance.
Females tend to do more homework than males.
The older you are the more homework you do.
Negative rho: if your friends do homework, you are less likely to do it!

Impacts:

```{r }
set.seed(123456)
im<-impacts(slm, listw=friends, zstats=TRUE, R=200)
sums<-summary(im,  zstats=T)
data.frame(sums$res)
data.frame(sums$pzmat)
```

Direct effects tend to be larger than indirect ones and of the reversed sign.
Direct effect: if you are a girl, you tend to do more homework than if you are a boy.
This impact includes the effect of feedback loops where you affect your friend and your friend influences you.
So if you are a girl, this will affect your homework, then will pass through the homework of friends, and then back to your homework.
Indirect effect: if your friend is a girl, you will also tend to do less homework, taking into account the feedback loop where your friend influences their friends etc.
Rho is negative: if your friends do more homework, you will do less homework.
The difference between the impacts and the point estimates that we had before calculating impacts is that point estimates do not take feedback loops into account.
Thus, interpret the impacts when having an autoregressive term.

## SEM and SAC

Spatial error model:

```{r }
sem <- errorsarlm(acthomew ~ sex + age, data = friendsnodelist, friends, zero.policy = TRUE, Durbin=FALSE)
summary(sem)
```

#Here impacts are the same as point estimates as there are no autoregressive terms.
#The lambda is negative: it means that unobserved factors of the friends are reducing your homework.

Spatial autoregressive combined model:

```{r }
sac <- sacsarlm(acthomew ~ sex + age, data = friendsnodelist, friends, zero.policy = TRUE, Durbin=FALSE)
summary(sac)
```

Also calculate impacts given that there is an autoregressive term. Same interpretation as above.

## SDM, SDEM, and GNM

Spatial Durbin model:

```{r }
sdm <- lagsarlm(acthomew ~ sex + age, data = friendsnodelist, friends, zero.policy = TRUE, Durbin=TRUE)
summary(sdm)
```

Also calculate impacts given that there is an autoregressive term. Same interpretation as above.
Note that the lag terms are the point estimates of the friends, but without taking into account the feedback loops.

Spatial Durbin error model:

```{r }
sdem <- errorsarlm(acthomew ~ sex + age, data = friendsnodelist, friends, zero.policy = TRUE, Durbin=TRUE)
summary(sdem)
```

No need to calculate impacts here as there are no autoregressive terms.

General nested model:

```{r }
gnm <- sacsarlm(acthomew ~ sex + age, data = friendsnodelist, friends, zero.policy = TRUE, Durbin=TRUE)
summary(gnm)
```

Also calculate impacts here. Same interpretation as in the other models.

## Conclusion

Think about it!